---
title: "Exponential_Smoothing"
author: "S. Sudhir"
date: "2023-08-16"
output: 
  html_document: 
    toc: yes
    df_print: tibble
---

# Exponential smoothing

Consider the `nhtemp` dataset which holds yearly average measurements of temperature for New Hampshire, from 1912 to 1971

```{r}
library(ggplot2)
require(datasets)
df1 = data.frame(year = 1912:1971, avg.temp = nhtemp)
```

```{r, echo = F}
df1
```

```{r, message = F, echo = F}
ggplot(data = df1, aes(x = year, y = avg.temp)) +
  
  ## points of original avg.temp vs year
  geom_point()+
  
  ## Plotting original avg.temp
  geom_line(data = df1, 
            aes(x = year, y = avg.temp, 
            color = "Original Time Series")) +
  
  ## Adding Legend
  scale_color_manual("Legend", breaks = c("Original Time Series"),
                                values = c("steelblue")) +
  
  ## Creating Titles for main, x-axis, and y-axis
  ggtitle('Yearly Average Measurements of Temperature for New Hampshire') + 
  xlab('Year') + 
  ylab('Average Temperature') + 
  
  ## Setting theme
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))
```

We want to fit an exponential smoothing model to this data such that
$\hat{Y}_1 = Y_1$ and, for $i = 2, 3, \ldots, n$,

$$\hat{Y}_i = \beta Y_{i-1} + (1-\beta) \hat{Y}_{i-1}$$

where $\beta$ is a constant between 0 and 1.

```{r}
## Forecast function as given above
## Estimated parameter: beta
## Fixed and observable parameter: y
forecast = function(beta, y)
{
  yhat = c()
  yhat[1] = y[1] # initializing yhat[1]
  
  for (i in 2:length(y))
  {
    yhat[i] = beta * y[i-1] + (1 - beta) * yhat[i-1] # recursion for yhat[2] on wards
  }
  return (yhat)
}
```

We will choose the parameter estimate $\hat{\beta}$ that minimizes the mean forecast error

$$FE(\beta) = \frac{1}{n} \sum_{i=2}^n \left( Y_i - \hat{Y}_i \right)^2$$

```{r}
## Parameter beta estimate which minimizes mean forecast error
forecast.error = function(beta, y)
{
  yhat = forecast(beta, y)
  return (1/length(y) * sum((y - yhat)^2))
}
```

The derivatives of this function are rather complicated (notice that $\hat{Y}_i$ is a function of $\beta$), so let's use a derivative-free method based on the function `optimize()`.

## Using `optimize()` on the interval $[0,1]$, to find the value of $\beta$ that produces the minimum forecast error.

```{r}
## Finding parameter estimator that minimizes forecast error
FE_opt = optimize(f = forecast.error,
                  interval = c(0,1),
                  tol = 0.0001,
                  y = df1$avg.temp)
FE_opt
```

```{r, echo = F}
paste0("The value of beta that minimizes forecast error is ", 
       FE_opt$minimum,
       " with the value at ", 
       FE_opt$objective)
```

```{r, echo=F}
## Merging original data with forecast using optimum beta found using optimize()
opt.forecast = forecast(FE_opt$minimum, y = df1$avg.temp)
df2 = data.frame(df1, opt.forecast)
```

```{r, message = F, echo=F}
ggplot(data = df2, aes(x = year, y = avg.temp)) +
  
  ## points of original avg.temp vs year
  geom_point()+
  
  ## Plotting original avg.temp
  geom_line(data = df2, 
            aes(x = year, y = avg.temp, 
            color = "Original Time Series")) +
  
  ## Plotting exponential smoothing
  geom_line(data = df2, 
            aes(x = year, y = opt.forecast, 
            color = "Exponential Smoothing BFGS")) +
  
  ## Adding Legend
  scale_color_manual("Legend", breaks = c("Original Time Series",
                                          "Exponential Smoothing BFGS"),
                                values = c("steelblue", 
                                           "coral")) +
  
  ## Creating Titles for main, x-axis, and y-axis
  ggtitle('Yearly Average Measurements of Temperature for New Hampshire') + 
  xlab('Year') + 
  ylab('Average Temperature') + 
  
  ## Setting theme
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))
```

## Reproducing the previous plot, including some other levels of smoothing, with $\beta=0.05$ and $\beta=0.75$

```{r, echo = F}
## Merging original data with forecast using beta = 0.05 and beta = 0.75
forecast_b1 = forecast(beta = 0.05, y = df1$avg.temp)
forecast_b2 = forecast(beta = 0.75, y = df1$avg.temp)

df3 = data.frame(df2, forecast_b1, forecast_b2)
```

```{r, message = F, echo = F}
ggplot(data = df3, aes(x = year, y = avg.temp)) +
  geom_point()+
  
  ## Original Time Series
  geom_line(data = df3, 
            aes(x = year, 
                y = avg.temp, 
                color = "Original Time Series",
                linetype = "Original Time Series")) +
  
  ## Exponential Smoothing BFGS
  geom_line(data = df3, 
            aes(x = year, 
                y = opt.forecast, 
                color = "Exponential Smoothing BFGS",
                linetype = "Exponential Smoothing BFGS")) +
  
  ## Exponential Smoothing beta: 0.1, color: lightpink, linetype: dashed
  geom_line(data = df3, 
            aes(x = year, 
                y = forecast_b1, 
                color = "Exponential Smoothing beta = 0.1",
                linetype = "Exponential Smoothing beta = 0.1")) +
  
  ## Exponential Smoothing beta: 0.9, color: seagreen, linetype: dashed
  geom_line(data = df3, 
            aes(x = year, 
                y = forecast_b2, 
                color = "Exponential Smoothing beta = 0.9",
                linetype = "Exponential Smoothing beta = 0.9")) +
  
  ## Adding Legend
  scale_color_manual("Legend", breaks = c("Original Time Series",
                                          "Exponential Smoothing BFGS",
                                          "Exponential Smoothing beta = 0.1",
                                          "Exponential Smoothing beta = 0.9"),
                                values = c("Original Time Series" = "steelblue", 
                                           "Exponential Smoothing BFGS" = "coral",
                                           "Exponential Smoothing beta = 0.1" = "lightpink",
                                           "Exponential Smoothing beta = 0.9" = "seagreen")) +
  
  scale_linetype_manual("Legend", breaks = c("Original Time Series",
                                          "Exponential Smoothing BFGS",
                                          "Exponential Smoothing beta = 0.1",
                                          "Exponential Smoothing beta = 0.9"),
                                values = c("Original Time Series" = "solid", 
                                           "Exponential Smoothing BFGS" = "solid",
                                           "Exponential Smoothing beta = 0.1" = "dashed",
                                           "Exponential Smoothing beta = 0.9" = "dashed")) +
  
  ## Creating Titles for main, x-axis, and y-axis
  ggtitle('Yearly Average Measurements of Temperature for New Hampshire') + 
  xlab('Year') + 
  ylab('Average Temperature') + 
  
  ## Setting theme
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))
```