---
title: "Example of single variable one-dimensional optimization"
author: "S. Sudhir"
date: "2023-08-17"
output: html_document
---

```{r, setup = TRUE, message = F, include = F}
knitr::opts_chunk$set(fig.width = 9, fig.height = 6) 
```

# One-dimensional optimization

An object subject to linear drag due to air resistance is projected upward at a specified velocity.
Its altitude $z$ is modeled as a function of time $t$ as:

$$z(t) = z_0 + \frac{m}{c}\bigg(v_0 + \frac{mg}{c}\bigg)\bigg(1 - e^{\frac{-c}{m}t}\bigg) - \frac{mg}{c}t$$

where:

$z:$ altitude (in m) above Earth’s surface (which has $z = 0$)

$z_0:$ initial altitude (in m)

$m:$ mass (in kg)

$c:$ Linear Drag Coefficient (in kg/s)

$v_0:$ initial velocity (in m/s) where positive velocity is up

$t:$ time (in s)

```{r}
# Create height function with inputs for:

# time: time in seconds
# alt.init: initial altitude in meters
# mass: mass of object in kg
# vel.init: initial velocity in m/s
# drag: drag coefficients in kg/s
# g: gravity default to 9.81 m/s2

altitude.fun <- function(time, alt.init, mass, drag, vel.init, g)
{
  altitude = alt.init + 
             ( (mass/drag) * (vel.init + ((mass*g)/drag)) * (1 - exp((-drag/mass)*time)) ) - 
             ( (mass*g/drag)*time )
  
  altitude.out = ifelse(altitude < 0, 0, altitude) # if object hits ground set value to 0
  return(altitude.out)
}
```

## Graph of function

```{r, message = F, echo = F}
library(ggplot2) # Library used to graph 2D functions

alt.init = 100 # Initial Height
vel.init = 55 # Initial Velocity
mass = 80 # Mass
drag = 15 # Linear Drag Coefficient
g = 9.81 # Force of Gravity
time = seq(from = 0, to = 15, by = .1) # Time from 0 to 15 seconds

# Running altitude.fun for time 0 to 15 seconds
y = altitude.fun(time, alt.init, mass, drag, vel.init, g)

# Plotting Function from above for time 0 seconds to 15 seconds
ggplot(data.frame(y, time), aes(x = time, y = y)) + 
  geom_point() + ggtitle('Altitude vs. Time') + 
  xlab('Time (in s)') + 
  ylab('Altitude (in m)') + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black"))
```

## Time when object strikes the ground

```{r}
# Running For Loop to find the first instance of time when ball hits the ground, 
# and then optimizing the function from 0 to that time t

for (t in time)
{
  if (altitude.fun(t, alt.init, mass, drag, vel.init, g) == 0)
  {
    min_t = optimize(f = altitude.fun, 
                          interval = c(0,t), 
                          alt.init = 100, 
                          mass = 80, 
                          drag = 15, 
                          vel.init = 55, 
                          g = 9.81, 
                          tol = 0.01)
    break
  }
}

print(paste0("Time it takes for the object to reach the ground is: ", 
             sprintf(min_t$minimum, fmt = '%#.3f')," seconds"))
```

```{r, echo = F}
# Plotting Function with minimum height indicator
ggplot(data.frame(y, time), aes(x = time, y = y)) + 
  geom_point() + ggtitle('Altitude vs. Time') + 
  xlab('Time (in s)') + 
  ylab('Altitude (in m)') + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  geom_vline(xintercept = min_t$minimum, 
             linetype = 'dashed', 
             color = 'coral', 
             linewidth = 0.8) +
  geom_point(aes(x=min_t$minimum,
                 y=min_t$objective), 
             colour="coral2")  + 
  geom_text(aes(label = round(min_t$minimum, 3), 
                x = min_t$minimum, 
                y = -Inf), 
            vjust = 1.5, col = 'coral') + 
  coord_cartesian(clip = 'off') 
```

## Find the object's maximum height

```{r}
# Using optimize to find both time and height where function reaches maximum
max_t = optimize(f = altitude.fun, 
                          interval = c(0,15), 
                          alt.init = 100, 
                          mass = 80, 
                          drag = 15, 
                          vel.init = 55, 
                          g = 9.81, 
                          tol = 0.01,
                          maximum = TRUE)

print(paste0("The maximum height the object reaches is : ", 
             sprintf(max_t$objective, fmt = '%#.3f')," meters")) 
```

```{r, echo = F}
# plotting graph with horizontal indicator
ggplot(data.frame(y, time), aes(x = time, y = y)) + 
  geom_point() + ggtitle('Altitude vs. Time') + 
  xlab('Time (in s)') + 
  ylab('Altitude (in m)') + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  geom_hline(yintercept = max_t$objective, 
             linetype = 'dashed', 
             color = 'coral', 
             linewidth = 0.8) +
  geom_point(aes(x=max_t$maximum, 
                 y=max_t$objective), 
             colour="coral2") + 
  geom_text(aes(label = round(max_t$objective, 3), 
                x = 0, 
                y = max_t$objective), 
            vjust = 1.5, 
            col = 'coral') + 
  coord_cartesian(clip = 'off') 
```

## Find the time at which the object reaches its maximum height

```{r, echo = F}
print(paste0("Time it takes for the object to reach the maximum height is: ", 
             sprintf(max_t$maximum, fmt = '%#.3f')," seconds"))

# Plotting graph with vertical indicator
ggplot(data.frame(y, time), aes(x = time, y = y)) + 
  geom_point() + ggtitle('Altitude vs. Time') + 
  xlab('Time (in s)') + 
  ylab('Altitude (in m)') + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) +
  geom_vline(xintercept = max_t$maximum, 
             linetype = 'dashed',
             color = 'coral', 
             linewidth = 0.8) +
  geom_point(aes(x=max_t$maximum,
                 y=max_t$objective), 
             colour="coral2") + 
  geom_text(aes(label = round(max_t$maximum, 3), 
                x = max_t$maximum, y = -Inf), 
            vjust = 1.5, col = 'coral') + 
  coord_cartesian(clip = 'off') 
```
